---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Set `echo=TRUE` as a global option to show code of every chunk in the document:
```{r setoptions, echo=TRUE}
library(knitr)
opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

If not already done, decompress the data archive. Then load it
```{r}
if(!file.exists("activity.csv")) unzip("activity.zip", overwrite = F)
activity <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

Calculate the total number of steps taken per day and plot histogram
```{r}
suppressMessages(library(dplyr))
activity.per.day <- group_by(activity, date) %>% summarize(steps.per.day = sum(steps, na.rm = T))
hist(activity.per.day$steps.per.day, xlab = "Steps", main = "Histogram of the total number of steps taken each day")
```

Mean and median of the total number of steps taken per day
```{r}
mean(activity.per.day$steps.per.day, na.rm = T)
median(activity.per.day$steps.per.day, na.rm = T)
```

## What is the average daily activity pattern?

```{r}
avg.steps <- group_by(activity, interval) %>% summarise(avg.steps = mean(steps, na.rm = T))
plot(avg.steps$avg.steps ~ avg.steps$interval, type = "l", main = "Average number of steps taken in day intervals")
```

The 5-minute interval which contains the maximum number of steps, on average across all the days in the dataset, is 
```{r}
as.character(avg.steps[avg.steps$avg.steps == max(avg.steps$avg.steps), "interval"])
```


## Imputing missing values

Total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
sum(!complete.cases(activity))
```

`f.activity` is a new dataset in which we fill `NA` values with the mean across all days value for that interval
```{r}
nas <- which(is.na(activity$steps))
f.activity <- activity
f.activity[nas, "steps"] <- avg.steps[avg.steps$interval %in% activity[nas, "interval"], "avg.steps"]
```

Make a histogram of the total number of steps taken each day
```{r}
tot.per.day <- group_by(f.activity, date) %>% summarise(tot.steps = sum(steps))
hist(tot.per.day$tot.steps, main = "Total number of steps taken each day")
```

Calculate and report the mean and median total number of steps taken per day.
```{r results='asis'}
library(xtable)
df <- data.frame(Mean = c(mean(activity.per.day$steps.per.day, na.rm = T), mean(tot.per.day$tot.steps)), 
                  Median = c(median(activity.per.day$steps.per.day, na.rm = T), median(tot.per.day$tot.steps)) 
                  )
rownames(df) <- c("Original data", "Filled data")
print(xtable(df), type = "html")
```

Mean and median calculated in the NA-filled data are greater than the original data mean and median.

## Are there differences in activity patterns between weekdays and weekends?

```{r}
f.activity <- mutate(f.activity, 
                     weekday = factor(ifelse(weekdays(as.Date(date)) %in% c("sabato", "domenica"), "weekend", "weekday")))
mean.f.activity <- group_by(f.activity, weekday, interval) %>% summarise(mean.steps = mean(steps))
library(ggplot2)
qplot(x = interval, y = mean.steps, data = mean.f.activity, geom = "line") + 
  facet_wrap(facets = ~ weekday, nrow = 2) + ylab("Number of steps")
```

